\chapter{The giant impact}
\label{ch05}
\graphicspath{{./05figs/}}

\section{Introduction}
present canonical model\\
difficulties\\
historical break-down\\
disagreement between different codes\\

\citep{Canup:2010p3713}

refer to disk masses in chapter 3\\


\section{mixing in miscible SPH vs. standard SPH}
% compare miscible with standard SPH


\section{Alternative giant impact model}
% new simulations

% additional figures
%TODOPLOT: disk thermal profiles (temp as color, ecc vs. rad)
%TODOPLOT (, ecc vs. rad)

Making the Moon from the Earths mantle




Andreas Reufer, Matthias M. M. Meier, Willy Benz and Rainer Wieler
\emph{submitted to Nature letters}

\textbf{The formation of the moon from a disk formed by a collision between the proto-Earth and a roughly Mars-sized impactor is widely accepted today. This „giant impact“ hypothesis[1] explains the high angular momentum of the Earth-Moon system and the moon's deficiency in iron. Hydrodynamical simulations have shown that a differentiated impactor with a chondritic Si/Fe ratio hitting the proto-Earth at mutual escape velocity and a grazing impact angle of 45° can lead to a proto-lunar disk sufficiently massive to later form the Moon [2]. Simulations of this “canonical” case consistently show that only about 20\% of the mass of the (later) Moon is derived from the proto-Earth. Thus, it is surprising that lunar rocks show a strong elemental and isotopic similarity to the Earth's mantle, considering the elemental and isotopic heterogeneity of different solar system materials. Equilibration of a partially molten and partially vaporized post-impact disk with a terrestrial rock vapour atmosphere has been put forward[3] to explain this similarity. However, so far no complete, quantitative model of this equilibration exists[4]. Here we present simulations of a new class of collisions with higher impact velocities and a steeper impact angle. They result in an iron-poor Moon with the angular momentum of the Earth-Moon system and a considerably higher fraction (50-80\%) of proto-Earth material in the moon-forming disk. Shock heating during impact is much stronger than in the canonical case and the proto-lunar disk material is initially vaporized to a larger extent, possibly facilitating additional re-equilibration of the post-impact disk with the terrestrial mantle. We also investigate the influence of impactor composition (in silicon, iron and ice) on the outcome of the collision.}

The collisional parameters for the giant impact considered so far have been limited to low-velocity collisions at, or only slightly above mutual escape velocity between target and impactor. The Earth-Moon system did not loose more than 10\% of its initial angular momentum (1 LE-M) between the giant impact and today[5]. In low-velocity collisions, very little mass and therefore also very little angular momentum is lost. Hence if impact velocity and angular momentum of the collision are fixed values, for a given impactor and target composition only one degree of freedom remains in the form of the product of the sine of impact angle and the impactor mass. Previous work therefore focussed on finding the optimum mass ratio between impactor and target. Most recent work suggests a mass ratio of 9:1 with a total mass of 1.05 ME [6]. Both the impactor and the target are assumed to be differentiated bodies with a 30wt\% iron core and a 70wt\% silicate mantle. In all these low-velocity collisions the impactor loses kinetic energy whilst grazing past the target and is then dispersed into a disk around the target. As a result, the proto-lunar disk is composed mainly of impactor material. The fraction of target silicate to total silicate material in the disk

\begin{equation}
f_T = (\Mtar / M_{tot})_{disk}
\end{equation}


is only around 20\%. If we define a similar target silicate material fraction for the post-impact Earth, we can deduce a depletion factor
  
$\delta f_{T}$ = (Mtarg / Mtot)disk / (Mtarg / Mtot)post-impact Earth – 1 (2)

which directly reflects the compositional similarity between the proto-lunar disk and the silicate portion of the Earth. This is important because the disk does not have to be derived completely from terrestrial (or target) material to match observations. It suffices that  the impactor contributes approximately the same fraction of material to the disk and the post-impact Earth. Therefore, it is this value $\delta f_{T}$ that has to approach ~0\% (within the respective errors) to match geochemical observations.

These geochemical observations indicate that the composition of lunar rocks is best understood if their source material is either derived mainly from the Earth's mantle, or if the lunar accretion disk was thoroughly re-equilibrated with Earth's mantle in the aftermath of the collision. The oxygen isotope compositions of the Earth and the Moon are so close to each other (Δ17O  <= 5 ppm [7]) that, e.g, for an impactor that is as different in 17O as Mars (Δ17O ~320 ppm), $\delta f_{T}$ can be no larger than about ~5\% (see Table 1 for the different values of $\delta f_{T}$ required by the different isotopic systems and impactor compositions). However, as there are at least two chondrite groups (enstatite and CI chondrites) that plot on the terrestrial mass-dependant fraction line in the oxygen three isotope diagram[8], it is possible that the terrestrial composition is representative of a larger solar system reservoir, from which also the impactor could have been derived. A further constraint is the Si isotopic composition of Earth and Moon, which are also identical within error, but distinct from all chondrite groups, Vesta and Mars [9], [10]. This was proposed to be the result of Si partitioning into the Earth's iron core, a process that is inefficient on planets that have less than ~15-20\% of  Earth's mass[11]. Therefore, this distinct isotopic signature cannot have formed independently on the moon, or the impactor (unless the impactor was on the very heavy end of the mass range[6], again linking the Moon to the terrestrial mantle. The lunar and terrestrial Mg/Cr ratios are similar (~100 and ~87) but differ clearly from the chondritic value (~36), which has also been explained by incorporation of Cr into Earth's core, a process that is inefficient on planets of the size of Mars [12]. In addition, the 53Cr/52Cr ratio is identical within error for Earth and Moon, although it varies with heliocentric distance[13]. Finally, Earth and Moon have a similar Hf/W-ratio and essentially identical ε182W values [14]. While in giant impact simulations both bodies may end up with an identical ε182W value, in no case simulated by [15] were the observed Hf/W-ratios simultaneously matched. From all these observations, it was concluded15], [4] that either the moon had to be derived predominantly from the Earth's mantle, or that the post-impact disk and the Earth's mantle must have re-equilibrated. So far, the first of these two possible solutions was incompatible with the results of hydrodynamical simulations. In the following paragraphs, we present a new series of simulations, where for the first time a significantly higher fraction of the lunar material is derived from the Earth's mantle.

The collisions presented here fall into the broad regime of slow hit-and-run collisions[16], which has never before been considered for the giant impact. Because of the higher impact velocities (1.2-4 vesc) in this type of collisions, substantial mass and angular momentum can be lost in the process. Therefore the initial angular momentum is less constrained and can be considerably higher than 1 LE-M. Table 1 in the supplementary online material shows a selection of around 60 simulations of this type performed by us. The higher impact velocities of these collisions are also encouraged by current models of terrestrial planet formation [17]. 

We used a standard SPH code with self-gravity and the ANEOS equation of state [18] for iron and water ice and M-ANEOS [19] for silicate. All simulations used a resolution of around 500'000 particles and were performed in a similar manner as in previous work[20], [2], [5], [6]. 

A reference case (cA08) uses initial parameters and conditions comparable to those used in the canonical case[6] and successfully reproduces an iron-depleted proto-lunar disk massive enough to form a Moon.  For each of our simulations, the Moon mass is calculated using the disk mass and the specific angular momentum[21]. We found collisions with an angle of 30-40° and 1.2 – 1.3 vesc are successful to put target material into orbit, when using differentiated impactors with chondritic Si/Fe composition (30wt\% Fe, 70wt\% SiO2) and masses between 0.15-0.20 ME. Some cases in this regime show an iron excess of > 5w\% in the proto-lunar disk and are rejected, as in previous work[6]. The best chondritic case matching all constraints is obtained using an impact angle of 35° and a velocity of 1.20 vesc,, where 56\% of the material in the silicate part of the disk originates from the target and $\delta f_{T}$ reduces to -35\% compared to -66\% in the canonical reference case. Figure 1 shows four consecutive snapshots of such a high velocity collision.

The origin of the matter ending up in the proto-lunar disk is mainly defined by the geometry of the collision and is determined during the very early phase when the impactor is accelerating the target material. This can be seen in Figure 2 where the particles which later end up in the disk are highlighted in bright colour. In the canonical case, the impactor grazes around the target's mantle and is deformed. Due to the low impact velocity, material supposed to end up in orbit around the Earth must not be decelerated too strongly in order to retain enough energy to stay in orbit. In the canonical case, this is only the case for the parts of the impactor mantle most distant to the point of impact and some minor part of the target's mantle. But if impact velocity is increased from 1.00 (cA08) to 1.30 vesc (cC01), parts from deeper within the target mantle receive the right amount of energy for orbit insertion. The outer regions of the target mantle receive too much energy and leave the system. This material removes mass and angular momentum from the system. While in case cB04 only ~ 10\% of the initial angular momentum is removed, ~ 45\% are removed in case cC06 (see table 1 in SOM).

If the collision geometry predominantly determines the fraction of target material in the proto-lunar disk, altering the size of the impactor by density changes should also change the target material fraction in the disk. A denser impactor with the same mass delivers the same momentum, while reducing “spill-over” of impactor material as it can be seen in figure 2. High density, iron-rich impactors might be a product of composition changing hit-and-run collisions[22]. We investigated this scenario with 50\%wt and 70wt\% iron core fraction impactors. With a 0.2ME impactor at 30° impact angle and 1.30 vesc impact velocity the target material fraction fT increases from 57\% ($\delta f_{T}$ = -34\%) in the chondritic case (cC01), to 64\% ($\delta f_{T}$ = -28\%) in the 50wt\% iron core case (fA01) up to 75\% ($\delta f_{T}$ = -19\%) in the 70wt\% iron core case (fB06). However with the target material fraction, also the bound angular momentum and iron content of the disk increase to unrealistic values. Apparently, reducing the “spill-over” also reduces the lost mass and therefore lost angular momentum. 

We also looked into less dense but still fully differentiated impactors with a composition of 50wt\% ice, 35wt\% silicate and 15wt\% iron, typical for small bodies accreted in regions of the solar system beyond the snow-line. Such a volatile-rich impactor scenario with an oxidized mantle is motivated  by terrestrial Pd-Ag and Cr, Sr isotope geochemistry data, suggesting the very late accretion of 10-20\% of CI-chondrite like material[23]. In this scenario, the efficiency of putting material into orbit is reduced compared to cases with denser impactors, though the fact that there is less impactor silicate material available to end up in the disk actually raises the target material fraction to 81\%, and $\delta f_{T}$ to -10\% in the case of an 0.2ME impactor hitting at 30° impact angle and 1.30 vesc impact velocity. However the resulting silicate portion of the disk is not massive enough (0.73 ML) to later form the Moon.  

The target material depletions yielded by the different models are compatible with geochemical constraints, with the possible exception of oxygen (see table 1). They can also explain the difference in FeO content observed in the mantles of Earth and Moon (8wt\% and 13wt\% FeO, respectively[12]) by simple mixture of an impactor mantle with a certain FeO content and a proto-Earth with a FeO content of ~8wt\% (assuming this is the equilibrium value determined by the oxygen fugacity and core size[24]). A typical $\delta f_{T}$ of -35\% for a 0.2 ME impactor then results in an impactor mantle FeO content of ~18\%, identical with Mars (see Figure 1 in SOM).

While the high velocity collision scenarios suit the dynamical and geochemical post-impact constraints as required for Moon formation, they differ considerably from the canonical case in terms of thermal history. The higher impact velocity leads to stronger shocks and therefore to more intense impact heating. Figure 3 shows cuts through the post-impact Earth after ~ 58h with colour-coded temperature. In the canonical case (Figure 3, left plot), only a thin blanket of material raining back from the disk experienced strong heating, while most of the mantle is only moderately heated above the initial average temperature of 2000K used in the simulation. In the high velocity case the shock is stronger and heats a substantial part of the mantle to temperatures above 10'000K. Beside possible implications for the early history of the Earth, this also constitutes a different starting point for the re-equilibration model[3], where the mantle of the post-impact Earth has to be fully convective in order to allow full isotopic re-equilibration with the post-impact disk. While in the canonical model the thin hot blanket may inhibit convection at early stages, the much thicker hot layer which extends deep into the mantle in our model probably simplifies exchange of material between the post-impact Earth and the hot silicate vapor atmosphere. Our results not only relax the required efficiency of re-equilibration by starting with a disk with a higher target material fraction, but the resulting thermodynamical state might actually create a more favourable starting point for re-equilibration.

Another key difference to the canonical model is the dynamical state of the proto-lunar disk. In both models, material is ejected on ballistic trajectories after impact and forms an arm structure. Angular momentum is transferred from the inner to the outer parts, thus circularising the orbit of the outer parts of the arm while the inner arm re-impacts the Earth. In the high velocity cases, this arm structure persists for a smaller time compared to the canonical case, transferring less angular momentum and leaving the material on more eccentric orbits. The dynamical evolution of such an eccentric disk and the consequences for Moon formation have yet to be studied.

In summary, our new approach of a high velocity (1.30 vesc) hit-and-run collision can explain the isotopic similarity of the Earth and the Moon for all isotope systems except possibly oxygen (see table 1). This is a clear improvement over the canonical models[6], where none of the known isotopic similarities could be explained. Oxygen only then cannot be explained by our model if the impactor indeed was as different in oxygen isotopes as Mars. In that case, a re-equilibration model [3],[4] might explain the similarity. Our work then also constitutes a new, hotter starting point for this re-equlibration model. 

M.M.M. Meier and R. Wieler provided an overview of the geochemical constraints and motivated new simulations of the giant impact in regimes not yet considered before. A. Reufer and W. Benz performed the necessary simulations and the analysis of the data. All authors discussed the results and implications and commented on the manuscript at all stages. The authors have no competing interests as defined by Nature Publishing Group, or other interests that might be perceived to influence the results and/or discussion reported in this article.

We thank Maria Schönbächler, Kaveh Pahlevan and Vera Fernandes for discussions and Jay Melosh for providing us with the M-ANEOS package. Andreas Reufer and Matthias M. M. Meier were both supported by the Swiss National Science Foundation. All calculations were performed on the ISIS2 cluster at University of Bern.
