\cleardoublepage
\chapter*{Acknowledgements}

First and foremost, I would like to express my deepest gratitude to Prof. Willy Benz, my thesis advisor. He always provided me with an open ear alongside his busy schedule. During the past few years, he shared with me his broad and deep knowledge of planet formation and his long experience in the field of computational astrophysics. Several times during the PhD, he also gave me the opportunity to meet and exchange with colleagues and encouraged fruitful collaborations. I'm very grateful for his non-intrusive way of advising me and the freedom he granted me during my work.

Cordial thanks go to Prof. Erik Asphaug, who not only kindly agreed to referee this work, but also motivated me with his inspiring ideas, either in discussions or through his work. Reading his papers is a great and entertaining way of expanding one's vocabulary for non-native speakers of the english language.

Thanks go to Prof. Nicolas Thomas, Shane Byrne, Veronica Bray, Colin Dundas and Mindi Searls for broadening my horizon towards Mars and showing me that numerical models can actually be applied to the real world out there.

I would like to thank Matthias M. M. Meier and Prof. Rainer Wieler for an interesting and ongoing collaboration concerning the Moons formation and giving me insight into the complex world of geochemistry. A special thank goes to Matthias for teaching me some manners amongst geochemists by suggesting the use of the term \emph{silicates} instead of \emph{rocks}.

Many thanks also go the current and past members of our institute and especially our TAPS (Theoretical Astrophysics and Planetary Science) group, who provided an interesting and helpful work environment during all those years.

I would also like to all the non-scientific friends who kept me away from my computer and provided some distraction from time to time. Last but not least I would like to thank my parents Verena and Markus Reufer for supporting my wish to follow my interest. Thanks!

This work was kindly supported by the Swiss National Science foundation.
